Algebraic Principles for Rely-Guarantee Style Concurrency Verification Tools

### 2014-03-04 Paper/paper.tex
Added Aczel traces with process labels
Added environments

\documentclass{llncs}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{tikz}
\usepackage{graphicx,color}
\usepackage{multicol}
\usepackage{bussproofs}
\usepackage{float}
\usepackage{centernot}
\usepackage{marginnote}

@@

\begin{abstract}
  We provide simple equational principles for deriving
  rely-guarantee-style inference rules and refinement laws based on
  idempotent semirings. We link the algebraic layer with concrete
  models of programs based on languages and execution traces. We have
  implemented the approach in Isabelle/HOL as a lightweight
  concurrency verification tool that supports reasoning about the
  control and data flow of concurrent programs with shared variables
  at different levels of abstraction. This is illustrated on two
  simple verification examples.
\end{abstract}

@@

Second, we link our rely-guarantee algebras with a simple trace based
semantics which so far is restricted to finite executions and
disregards termination and synchronisation. Despite the simplicity of
this model, we demonstrate and evaluate our prototypical verification
tool implemented in Isabelle by verifying two simple examples from
the literature. Beyond that our approach provides a coherent framework
from which more complex and detailed models can be implemented in the
future.

@@

Taken together, our Isabelle implementation constitutes a tool
prototype for the verification and construction of concurrent
programs. We illustrate the tool with two simple examples from the
literature. The complete Isabelle code can be found
online\footnote{\url{www.dcs.shef.ac.uk/$\sim$alasdair/rg}}. A
previous Isabelle implementation of rely-guarantee reasoning is due to
Prensa Nieto~\cite{nieto_rely-guarantee_2003}. Our implementation
differs both by making the link between concrete programs and algebras
explicit, which increases modularity, and by allowing arbitrary nested parallelism.

@@

@@

It is well known that the inference rules of sequential Hoare logic
(except the assignment axiom) can be derived in expansions of Kleene
algebras. One approach is as
follows~\cite{moller_algebras_2006}. Suppose a suitable Boolean
algebra $B$ of \emph{tests} has been embedded into a Kleene algebra
$K$ such that $0$ and $1$ are the minimal and maximal element of $B$,
$+$ corresponds to join and $\cdot$ to meet. Complements $-$ are
defined only on $B$. Suppose further that a \emph{backward diamond
  operator} $\langle x|p$ has been defined for each $x\in K$ and $p\in
B$, which models the set of all states to which each terminating
execution of program $x$ may lead from states $q$. Finally suppose
that a \emph{forward box operator} $|x]p$ has been defined which
models the (largest) set of states from which every terminating
execution of $x$ must end in states $p$ and that boxes and diamonds
are adjoints of the Galois connection $\langle x|p \le q
\Leftrightarrow p \le |x]q$, for all $x\in K$ and $p,q\in B$. It is
then evident from the above explanations that validity of a Hoare
triple $\vdash \{p\}x\{q\}$ can be encoded as $\langle x|p \le q$ and
the weakest liberal precondition operator $\mathsf{wlp}(x,q)$ as
$|x]p$. Hence the relationship between the proof theory and the
semantics of Hoare logic is captured by the Galois connection $
\vdash\{p\}x\{q\} \Leftrightarrow p \le \mathsf{wlp}(x,q)$. It has
been shown that the relational semantics of sequential while-programs
can be encoded in these \emph{modal Kleene algebras} and that the
inference rules of Hoare logic can be
derived~\cite{moller_algebras_2006}.

@@

As an example we show the derivation of a generalised while
rule. Suppose $x\cdot t\cdot y\le x$. Then $x\cdot (t\cdot y)^\ast \le
x$ by the right induction axiom of Kleene algebra and therefore
$x\cdot (t\cdot y)^\ast t'\le x\cdot t'$ for arbitrary element $t'$ by
isotonicity of multiplication. This derives the while rule
\begin{equation*}
  \frac{\vdash \{x\cdot t\}y\{x\}}{\vdash \{x\}(t\cdot y)^\ast\{t'\cdot x\}}
\end{equation*}
for a generalised while loop $(t\cdot y)^\ast\cdot t'$, which
specialises to the conventional rule when $t$ and $t'$ are, in some
sense, complements.

@@

Jones quintuples can be encoded in this setting as
\begin{align}
r, g \vdash \{p\} x \{q\} \iff p\cdot(r\|x) \le q \land x \le g. \label{quin}
\end{align}
This means that program $x$ when constrained by a rely $r$, and
executed after $p$, behaves as $q$. Moreover, all behaviours of $x$
are included in its guarantee.

@@

\begin{figure}[hbt]
\centering
\begin{prooftree}
\RightLabel{\ Skip}
\AxiomC{$p\cdot r \le p$}
\UnaryInfC{$r, g \vdash \{p\}1\{p\}$}
\end{prooftree}

@@

\begin{prooftree}
\RightLabel{\ Parallel}
\AxiomC{$r_1, g_2 \vdash \{p_1\}x\{q_1\}$}
\AxiomC{$g_1 \le r_2$}
\AxiomC{$r_1, g_2 \vdash \{p_2\}y\{q_2\}$}
\AxiomC{$g_2 \le r_1$}
\QuaternaryInfC{$r_1 \sqcap r_2, g_1 \| g_2 \vdash \{p_1 \sqcap q_2\}x\|y\{q_1 \sqcap q_2\}$}
\end{prooftree}

@@

If we add residuals to our algebra quintuples can be encoded in the
following way, which is equivalent to the encoding in Equation
(\ref{quin}).
\begin{align}
r, g \vdash \{p\} c \{q\} \iff c \le r/(p \rightarrow q) \sqcap g \label{refine}.
\end{align}
This encoding allows us to think in terms of program refinement, as in~\cite{hayes_refining_2013}, since
$r/(p \rightarrow q) \sqcap g$ defines the weakest program that when
placed in parallel with interference from $r$, and guaranteeing
interference at most $g$, goes from $p$ to $q$---a generic
specification for a concurrent program.

@@

The obvious way of achieving this is to expand our rely-guarantee
algebra with an additional function $\pi : K \to K$ and redefining our
quintuples as,
\begin{align*}
r, g \vdash \{p\} x \{q\} \iff p\cdot (r\|c) \le_\pi q \land x \le g.
\end{align*}
Where $x \le_\pi y$ is $\pi(x) \le \pi(y)$. Since for any operator
$\bullet$ it is not required that
\begin{align*}
\pi(x) = \pi(y) \implies \pi(x \bullet z) = \pi(y \bullet z),
\end{align*}
we can break compositionality in just the right way, provided we chose
appropriate properties for $\pi$. These properties are extracted from
properties of the trace model, which will be explained in detail in
the next section. Many of those can be derived from the fact that, in
our model, $\pi = \lambda x.\; x \sqcap c$, where $c$ is healthiness
condition which filters out ill-defined traces. We do not list these
properties here. In addition $\pi$ must satisfy the properties
\begin{align}
x^\star &\le_\pi \pi(x)^\star, \label{con1}\\
x\cdot y &\le_\pi \pi(x)\cdot \pi(y), \label{con2}\\
z + x\cdot y \le_\pi y &\implies x^\star\cdot z \le_\pi y\label{con3},\\
z + y\cdot x \le_\pi y &\implies z\cdot x^\star \le_\pi y\label{con4}.
\end{align}

@@

We consider languages where the alphabet contains state pairs of the
form $(\sigma_1,\sigma_2) \in \Sigma^2$. A word in such a language is
\emph{consistent} if every such pair in a word has the same first
state as the previous transition's second state. For example,
$(\sigma_1,\sigma_2)(\sigma_2,\sigma_3)$ is consistent, while
$(\sigma_1,\sigma_2)(\sigma_3,\sigma_3)$ is inconsistent. Sets of
consistent words are essentially \emph{Aczel
  traces}~\cite{boer_formal_1999}, but lack the usual process
labels. We denote the set of all consistent words by $C$ and define
the function $\pi$ from the previous section as $\lambda X.\; X \cap
C$ in our model.

@@

  \path [rel] (s11) -- (s22);
  \path [rel] (s12) -- (s22);
  \path [rel] (s23) -- (s32);
\end{tikzpicture}
\end{center}
The language contains the following six words
\begin{align*}
(\sigma_1,\sigma_1)(\sigma_1,\sigma_2)(\sigma_2,\sigma_3), \qquad
(\sigma_1,\sigma_2)(\sigma_1,\sigma_2)(\sigma_2,\sigma_3),\\
(\sigma_2,\sigma_2)(\sigma_1,\sigma_2)(\sigma_2,\sigma_3), \qquad
(\sigma_1,\sigma_1)(\sigma_3,\sigma_2)(\sigma_2,\sigma_3),\\
(\sigma_1,\sigma_2)(\sigma_3,\sigma_2)(\sigma_2,\sigma_3), \qquad
(\sigma_2,\sigma_2)(\sigma_3,\sigma_2)(\sigma_2,\sigma_3),
\end{align*}
where only the first,
$(\sigma_1,\sigma_1)(\sigma_1,\sigma_2)(\sigma_2,\sigma_3)$ is
consistent. This word is highlighted with solid arrows in the
diagram above. Now if we shuffle the single state pair $(\sigma_2,
\sigma_3)$ into the above language, we might end up with a language as
below:

@@

With mumbling we now have that
\begin{align*}
\test(P \cap Q) \le_\pi \test(P); \test(Q)
\end{align*}
as the longer words in $\test(P);\test(Q)$ can be mumbled down into
the shorter words of $\test(P\cap Q)$, whereas stuttering gives us the opposite direction,
\begin{align*}
\langle \Id\rangle^\star \| (\test(P);\test(Q)) \le_\pi \langle \Id\rangle^\star \| \test(P \cap Q).
\end{align*}
We henceforth assume that all languages are
implicitly mumble closed.

Using tests, we can encode if statements and while loops
\begin{align*}
\mathsf{if }\;\; P \text{ \{ } X \text { \} \textsf{else} \{ } Y \text{ \}} &= \test(P); X + \test(- P); Y,\\
\mathsf{while}\;\; P \text{ \{ } X \text{ \}} &= (\test(P);X)^\star;\test(- P).
\end{align*}
Next, we define the operator $\edn(P)$ which contains all the
words which end in a state satisfying $P$. Some useful properties of $\edn$ include
\begin{gather*}
\edn(P); \test{Q} \le_\pi \edn(P \cap Q), \qquad
\test(P) \le \edn(Q),\\
\text{range}(\Id_P \circ R) \le P \implies \edn(P); \langle R\rangle^\star \le_\pi \edn(P).
\end{gather*}

In this model, assignment is defined as
\begin{align*}
  x := e \ =\ \bigcup v.\,\test\{\sigma.\,\eval(\sigma,e) = v\} \cdot x \leftarrow v
\end{align*}
where $x \leftarrow v$ denotes the atomic command which assigns the
value $v$ to $x$. The $\eval$ function evaluates an expression $e$ in
the state $\sigma$. Using this definition we derive the assignment
rule
\begin{align*}
&\unchanged(\mathsf{vars}(e)) \cap \preserves(P) \cap \preserves(P[x/e]),\\
&\unchanged(- \{x\})\\
&\vdash \{\edn(P)\}\ x := e\ \{\edn(P[x/e])\}.
\end{align*}

@@

To prove the correctness of FINDP, we must show that
\begin{align*}
\text{FINDP} \le_\pi \edn(\mathsf{leastP}(f)) + \edn(f = \mathsf{len}(\text{array})),
\end{align*}
where $\mathsf{leastP(f)}$ is the set of states where $f$ is the least
index satisfying $P$, and $f = len(array)$ is the set of states where
$f$ is the length of the array. In other words, either we find the
least element, or no elements in the array satisfy $P$, in which case
$f$ remains the same as the length of the array.

@@

We have introduced variants of semirings and Kleene algebras intended
to model rely-guarantee and interference based reasoning. We have
developed an interleaving model for these algebras which uses familiar
concepts from traces and language theory. This theory has been
implemented in the Isabelle/HOL theorem prover, providing a solid
mathematical basis on which to build a tool for mechanised refinement
and verification tasks. In line with this aim, we have applied
our formalisation to two simple examples.

@@

Algebra plays an important role in our development. First, it
allowed us to derive inference rules rapidly and with little proof
effort. Second, it yields an abstract layer at which many properties
that would be difficult to prove in concrete models can be verified
with relative ease by equational reasoning. Third, as pointed out in
Section \ref{sec:KA}, some fragments of the algebras considered are
decidable. Therefore, decision procedures for some aspects of
rely-guarantee reasoning can be implemented in interactive theorem
proving tools such as Isabelle.

>>>>

\documentclass{llncs}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{tikz}
\usepackage{graphicx,color}
\usepackage{multicol}
\usepackage{bussproofs}
\usepackage{float}
\usepackage{centernot}
\usepackage{marginnote}
\usepackage{hyperref}

@@

\begin{abstract}
  We provide simple equational principles for deriving
  rely-guarantee-style inference rules and refinement laws based on
  idempotent semirings. We link the algebraic layer with concrete
  models of programs based on languages and execution traces. We have
  implemented the approach in Isabelle/HOL as a lightweight
  concurrency verification tool that supports reasoning about the
  control and data flow of concurrent programs with shared variables
  at different levels of abstraction. This is illustrated on a simple
  verification example.
\end{abstract}

@@

Second, we link our rely-guarantee algebras with a simple trace based
semantics which so far is restricted to finite executions and
disregards termination and synchronisation. Despite the simplicity of
this model, we demonstrate and evaluate our prototypical verification
tool implemented in Isabelle by verifying a simple example from
the literature. Beyond that our approach provides a coherent framework
from which more complex and detailed models can be implemented in the
future.

@@

Taken together, our Isabelle implementation constitutes a tool
prototype for the verification and construction of concurrent
programs. We illustrate the tool with a simple example from the
literature. The complete Isabelle code can be found
online\footnote{\url{www.dcs.shef.ac.uk/~alasdair/rg}}. A
previous Isabelle implementation of rely-guarantee reasoning is due to
Prensa Nieto~\cite{nieto_rely-guarantee_2003}. Our implementation
differs both by making the link between concrete programs and algebras
explicit, which increases modularity, and by allowing arbitrary nested parallelism.

@@

A (unital) \emph{quantale} is a dioid based on a complete lattice
where the multiplication distributes over arbitrary suprema. Formally, it is a
structure $(S,\le,\cdot,1)$ such that $(S,\le)$ is a complete lattice,
$(S,\cdot,1)$ is a monoid and
\begin{align*}
x(\Sigma Y) = \Sigma\{xy|y\in Y\}, \qquad
(\Sigma X)y = \Sigma\{xy|x\in X\}.
\end{align*}
In a quantale the star is the sum of all powers $x^n$. Therefore, all
quantales are also Kleene algebras.

@@

It is well known that the inference rules of sequential Hoare logic
(except the assignment axiom) can be derived in expansions of Kleene
algebras. One approach is as
follows~\cite{moller_algebras_2006}. Suppose a suitable Boolean
algebra $B$ of \emph{tests} has been embedded into a Kleene algebra
$K$ such that $0$ and $1$ are the minimal and maximal element of $B$,
$+$ corresponds to join and $\cdot$ to meet. Complements $-$ are
defined only on $B$. Suppose further that a \emph{backward diamond
  operator} $\langle x|p$ has been defined for each $x\in K$ and $p\in
B$, which models the set of all states to which each terminating
execution of program $x$ may lead from states $p$. Finally suppose
that a \emph{forward box operator} $|x]p$ has been defined which
models the (largest) set of states from which every terminating
execution of $x$ must end in states $p$ and that boxes and diamonds
are adjoints of the Galois connection $\langle x|p \le q
\Leftrightarrow p \le |x]q$, for all $x\in K$ and $p,q\in B$. It is
then evident from the above explanations that validity of a Hoare
triple $\vdash \{p\}x\{q\}$ can be encoded as $\langle x|p \le q$ and
the weakest liberal precondition operator $\mathsf{wlp}(x,q)$ as
$|x]p$. Hence the relationship between the proof theory and the
semantics of Hoare logic is captured by the Galois connection $
\vdash\{p\}x\{q\} \Leftrightarrow p \le \mathsf{wlp}(x,q)$. It has
been shown that the relational semantics of sequential while-programs
can be encoded in these \emph{modal Kleene algebras} and that the
inference rules of Hoare logic can be
derived~\cite{moller_algebras_2006}.

@@

As an example we show the derivation of a generalised while
rule. Suppose $x\cdot t\cdot y\le x$. Then $x\cdot (t\cdot y)^\ast \le
x$ by the right induction axiom of Kleene algebra and therefore
$x\cdot (t\cdot y)^\ast\cdot t'\le x\cdot t'$ for arbitrary element $t'$ by
isotonicity of multiplication. This derives the while rule
\begin{equation*}
  \frac{\vdash \{x\cdot t\}y\{x\}}{\vdash \{x\}(t\cdot y)^\ast\cdot t'\{x\cdot t'\}}
\end{equation*}
for a generalised while loop $(t\cdot y)^\ast\cdot t'$, which
specialises to the conventional rule when $t$ and $t'$ are, in some
sense, complements.

@@

Jones quintuples can be encoded in this setting as
\begin{align}
r, g \vdash \{p\} x \{q\} \iff p\cdot(r\|x) \le q \land x \le g. \label{quin}
\end{align}
This means that program $x$ when constrained by a rely $r$, and
executed after $p$, behaves as $q$. Moreover, all behaviours of $x$
are included in its guarantee $q$. Note that this encoding is stronger
than in traditional rely-guarantee, as $x$ is required to
unconditionally implement $g$. The algebra could easily be extended
with an additional operator $f$ such that $f(r,x) \le q$ would encode
that $x$ implements $q$ only under interference of at most $r$. For
more complex examples than what we present in section
\ref{sec:Examples} such an encoding may prove neccessary.

@@

\begin{figure}[bth]
\centering
\begin{prooftree}
\RightLabel{\ Skip}
\AxiomC{$p\cdot r \le p$}
\UnaryInfC{$r, g \vdash \{p\}1\{p\}$}
\end{prooftree}

@@

\begin{prooftree}
\RightLabel{\ Parallel}
\AxiomC{$r_1, g_1 \vdash \{p_1\}x\{q_1\}$}
\AxiomC{$g_1 \le r_2$}
\AxiomC{$r_2, g_2 \vdash \{p_2\}y\{q_2\}$}
\AxiomC{$g_2 \le r_1$}
\QuaternaryInfC{$r_1 \sqcap r_2, g_1 \| g_2 \vdash \{p_1 \sqcap p_2\}x\|y\{q_1 \sqcap q_2\}$}
\end{prooftree}

@@

If we add residuals to our algebra quintuples can be encoded in the
following way, which is equivalent to the encoding in Equation
(\ref{quin}).
\begin{align}
r, g \vdash \{p\} x \{q\} \iff x \le r/(p \rightarrow q) \sqcap g \label{refine}.
\end{align}
This encoding allows us to think in terms of program refinement, as in~\cite{hayes_refining_2013}, since
$r/(p \rightarrow q) \sqcap g$ defines the weakest program that when
placed in parallel with interference from $r$, and guaranteeing
interference at most $g$, goes from $p$ to $q$---a generic
specification for a concurrent program.

@@

A way of achieving this is to expand our rely-guarantee
algebra with an additional function $\pi : K \to K$ and redefining our
quintuples as,
\begin{align*}
r, g \vdash \{p\} x \{q\} \iff p\cdot (r\|c) \le_\pi q \land x \le g.
\end{align*}
Where $x \le_\pi y$ is $\pi(x) \le \pi(y)$. Since for any operator
$\bullet$ it is not required that
\begin{align*}
\pi(x) = \pi(y) \implies \pi(x \bullet z) = \pi(y \bullet z),
\end{align*}
we can break compositionality in just the right way, provided we chose
appropriate properties for $\pi$. These properties are extracted from
properties of the trace model, which will be explained in detail in
the next section. Many of those can be derived from the fact that, in
our model, $\pi = \lambda x.\; x \sqcap c$, where $c$ is healthiness
condition which filters out ill-defined traces. We do not list these
properties here. In addition $\pi$ must satisfy the properties
\begin{align}
x^\star &\le_\pi \pi(x)^\star, \label{con1}\\
x\cdot y &\le_\pi \pi(x)\cdot \pi(y), \label{con2}\\
z + x\cdot y \le_\pi y &\implies x^\star\cdot z \le_\pi y\label{con3},\\
z + y\cdot x \le_\pi y &\implies z\cdot x^\star \le_\pi y\label{con4}.
\end{align}

@@

We consider languages where the alphabet contains state pairs of the
form $(\sigma_1,\sigma_2) \in \Sigma^2$. A word in such a language is
\emph{consistent} if every such pair in a word has the same first
state as the previous transition's second state. For example,
$(\sigma_1,\sigma_2)(\sigma_2,\sigma_3)$ is consistent, while
$(\sigma_1,\sigma_2)(\sigma_3,\sigma_3)$ is consistent only if $\sigma_2 = \sigma_3$. Sets of
consistent words are essentially \emph{Aczel
  traces}~\cite{boer_formal_1999}, but lack the usual process
labels. We denote the set of all consistent words by $C$ and define
the function $\pi$ from the previous section as $\lambda X.\; X \cap
C$ in our model.

@@

  \path [rel] (s11) -- (s22);
  \path [rel] (s12) -- (s22);
  \path [rel] (s23) -- (s32);
\end{tikzpicture}
\end{center}
The language contains the following six words
\begin{align*}
(\sigma_1,\sigma_1)(\sigma_1,\sigma_2)(\sigma_2,\sigma_3), \qquad
(\sigma_1,\sigma_2)(\sigma_1,\sigma_2)(\sigma_2,\sigma_3),\\
(\sigma_2,\sigma_2)(\sigma_1,\sigma_2)(\sigma_2,\sigma_3), \qquad
(\sigma_1,\sigma_1)(\sigma_3,\sigma_2)(\sigma_2,\sigma_3),\\
(\sigma_1,\sigma_2)(\sigma_3,\sigma_2)(\sigma_2,\sigma_3), \qquad
(\sigma_2,\sigma_2)(\sigma_3,\sigma_2)(\sigma_2,\sigma_3),
\end{align*}
where only the first,
$(\sigma_1,\sigma_1)(\sigma_1,\sigma_2)(\sigma_2,\sigma_3)$ is
consistent. This word is highlighted with solid arrows in the diagram
above. Now if we shuffle the single state pair $(\sigma_2, \sigma_3)$
into the above language, would end up with a language containing the
words represented in the diagram below:

@@

Assuming we apply mumbling to both sides of the following equation, we have that
\begin{align*}
\test(P \cap Q) \le_\pi \test(P); \test(Q)
\end{align*}
as the longer words in $\test(P);\test(Q)$ can be mumbled down into
the shorter words of $\test(P\cap Q)$, whereas stuttering gives us the opposite direction,
\begin{align*}
\langle \Id\rangle^\star \| (\test(P);\test(Q)) \le_\pi \langle \Id\rangle^\star \| \test(P \cap Q).
\end{align*}
We henceforth assume that all languages are
implicitly mumble closed.

Using tests, we can encode if statements and while loops
\begin{align*}
\mathsf{if }\;\; P \text{ \{ } X \text { \} \textsf{else} \{ } Y \text{ \}} &= \test(P); X + \test(- P); Y,\\
\mathsf{while}\;\; P \text{ \{ } X \text{ \}} &= (\test(P);X)^\star;\test(- P).
\end{align*}
Next, we define the operator $\edn(P)$ which contains all the
words which end in a state satisfying $P$. Some useful properties of $\edn$ include
\begin{gather*}
\edn(P); \test{Q} \le_\pi \edn(P \cap Q), \qquad
\test(P) \le \edn(P),\\
\text{range}(\Id_P \circ R) \le P \implies \edn(P); \langle R\rangle^\star \le_\pi \edn(P).
\end{gather*}

In this model, assignment is defined as
\begin{align*}
  x := e \ =\ \bigcup v.\,\test\{\sigma.\,\eval(\sigma,e) = v\} \cdot x \leftarrow v
\end{align*}
where $x \leftarrow v$ denotes the atomic command which assigns the
value $v$ to $x$. The $\eval$ function atomically evaluates an expression $e$ in
the state $\sigma$. Using this definition we derive the assignment
rule
\begin{align*}
&\unchanged(\mathsf{vars}(e)) \cap \preserves(P) \cap \preserves(P[x/e]),\\
&\unchanged(- \{x\})\\
&\vdash \{\edn(P)\}\ x := e\ \{\edn(P[x/e])\}.
\end{align*}

@@

To prove the correctness of FINDP, we must show that
\begin{align*}
\text{FINDP} \le_\pi \edn(\mathsf{leastP}(f)) + \edn(f = \mathsf{len}(\text{array})),
\end{align*}
where $\mathsf{leastP(f)}$ is the set of states where $f$ is the least
index satisfying $P$, and $f = len(array)$ is the set of states where
$f$ is the length of the array. In other words, either we find the
least element, or $f$ remains the same as the length of the array, in
which case no elements in the array satisfy $P$.

@@

We have introduced variants of semirings and Kleene algebras intended
to model rely-guarantee and interference based reasoning. We have
developed an interleaving model for these algebras which uses familiar
concepts from traces and language theory. This theory has been
implemented in the Isabelle/HOL theorem prover, providing a solid
mathematical basis on which to build a tool for mechanised refinement
and verification tasks. In line with this aim, we have applied
our formalisation to a simple example program.

@@

Algebra plays an important role in our development. First, it allowed
us to derive inference rules rapidly and with little proof
effort. Second, it yields an abstract layer at which many properties
that would be difficult to prove in concrete models can be verified
with relative ease by equational reasoning. Third, as pointed out in
Section \ref{sec:KA}, some fragments of the algebras considered are
decidable. Therefore, decision procedures for some aspects of
rely-guarantee reasoning can be implemented in interactive theorem
proving tools such as Isabelle. However, we have not yet
investigated the extent to which such decision procedures would benefit
our approach.
